Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-02 demystify.tex
updates

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should give details of the applicant's research background and achievements, all of which paint a compelling picture of the candidate to the adcom.

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little value and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Moreover, experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{Helping your LOR writer} As mentioned above, do not write your own letter. Instead, you can help your LOR writer by providing details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicit ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

Sometimes students would go to great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\paragraph{Waiving Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Reminding Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them about the deadline and ask them to submit their letters on time.  But do not send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

    \item \textbf{Not customizing the email}. Many students write a generic email and send it to multiple professors. Tthese emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Some students send to multiple emails using BCC or even CC (where we can see all addresses the email being sent to). This is a big turn-off for professors and will result in no reply or a very harsh reply on how unprofessional you are.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
\end{itemize}

>>>>

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should give details of the applicant's research background and achievements, all of which paint a compelling picture of the candidate to the adcom.
Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

\paragraph{Why Self-written Letters are Bad?} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \textbf{very little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{Why Generic Letters are Bad?} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak. This does not mean the professor is not good or does not care about you, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\paragraph{Why You Should Waive Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Why You Want to Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. One thing you can do to help your writer to write a strong, customized LoR for you is providing details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\paragraph{Why You Need to Remind Your Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.  Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

Thus you do want to send reminders because professors can be quite busy and might forget to submit their letters, especially when there is no explicit deadline. However, do not send too many reminders as that can be annoying to the writers.

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with PhD in CS in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

    \item \textbf{Not customizing the email}. These emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information. 2--3 paragraphs are enough.  Personally I also do not want to see course transcript or CV in the first email.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
\end{itemize}
